---
title: "Lab Report: Computer lab 1 block 1"
author: "Group_A10"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
---

# Assignment 1: Handwritten Digit Recognition with K-Nearest Neighbors

```{r, echo=FALSE}
library(kknn)       # For K-Nearest Neighbors
#Load and Preprocess Data
# Load dataset
file_path <- "data/optdigits.csv"
optdigits_data <- read.csv(file_path)
colnames(optdigits_data) <- c(paste0("pixel_", 1:64), "label")
optdigits_data$label <- as.factor(optdigits_data$label)


# Split into training and test sets
set.seed(123)
n <- nrow(optdigits_data)
# Generate indices
train_indices <- sample(1:n, size = 0.5 * n)  # 50% for training
remaining_indices <- setdiff(1:n, train_indices)  # Remaining 50%
test_indices <- sample(remaining_indices, size = 0.25 * n)  # 25% for testing
validation_indices <- setdiff(remaining_indices, test_indices)  # Remaining 25% for validation

# Subset the data
train_data <- optdigits_data[train_indices, ]
test_data <- optdigits_data[test_indices, ]
validation_data <- optdigits_data[validation_indices, ]

# Train the KNN classifier
knn_model <- kknn(label ~ ., train = train_data, test = train_data, k = 30, kernel = "rectangular")


# Predictions and performance
train_predictions <- fitted(knn_model)
conf_matrix_train <- table(Actual = train_data$label, Predicted = train_predictions)
# print("Confusion Matrix (Training):")
# print(conf_matrix_train)

# Predict on test data
test_predictions <- fitted(kknn(label ~ ., train = train_data, test = test_data, k = 30, kernel = "rectangular"))
conf_matrix_test <- table(Actual = test_data$label, Predicted = test_predictions)
# print("Confusion Matrix (Test):")
# print(conf_matrix_test)


# Misclassification error for training data
train_total <- sum(conf_matrix_train)  # Total instances in training data
train_correct <- sum(diag(conf_matrix_train))  # Correctly classified instances (sum of diagonal)
train_misclassification_error <- 1 - train_correct / train_total

# cat("Training Misclassification Error:", train_misclassification_error, "\n")

# Misclassification error for test data
test_total <- sum(conf_matrix_test)  # Total instances in training data
test_correct <- sum(diag(conf_matrix_test))  # Correctly classified instances (sum of diagonal)
test_misclassification_error <- 1 - test_correct / test_total

# cat("Test Misclassification Error:", test_misclassification_error, "\n")


# Get predicted probabilities for training data
train_probabilities <- knn_model$prob  # Matrix of probabilities
train_actual <- train_data$label  # Actual labels for training data

# Extract rows corresponding to digit "8" in the training data
digit_8_indices <- which(train_actual == "8")
digit_8_probs <- train_probabilities[digit_8_indices, ]  # Probabilities for digit "8"

digit_8_predicted_probs <- digit_8_probs[, "8"]

easiest_indices <- digit_8_indices[order(digit_8_predicted_probs, decreasing = TRUE)[1:2]]
hardest_indices <- digit_8_indices[order(digit_8_predicted_probs, decreasing = FALSE)[1:3]]

# Extract feature data (pixels) for easiest and hardest cases
easiest_features <- train_data[easiest_indices, 1:64]  # Features for easiest cases
hardest_features <- train_data[hardest_indices, 1:64]  # Features for hardest cases

# Reshape into 8x8 matrices
# Reshape and store as numeric matrices
easiest_matrices <- lapply(1:nrow(easiest_features), function(i) {
  matrix(as.numeric(easiest_features[i, ]), nrow = 8, byrow = TRUE)
})

hardest_matrices <- lapply(1:nrow(hardest_features), function(i) {
  matrix(as.numeric(hardest_features[i, ]), nrow = 8, byrow = TRUE)
})

# Visualize easiest cases
# heatmap(easiest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", main = paste("Easiest Case", i), xlab = "", ylab = "")
# heatmap(easiest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", main = paste("Easiest Case", i), xlab = "", ylab = "")
# 
# # Visualize hardest cases
# heatmap(hardest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", i), xlab = "", ylab = "")
# heatmap(hardest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", i), xlab = "", ylab = "")
# heatmap(hardest_matrices[[3]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", i), xlab = "", ylab = "")
# 
# 





# Initialize vectors to store errors
k_values <- 1:30  # Values of K to test
train_errors <- numeric(length(k_values))  # Training errors
validation_errors <- numeric(length(k_values))  # Validation errors

# Loop through each value of K
for (k in k_values) {
  # Train KNN model on training data
  knn_model <- kknn(label ~ ., train = train_data, test = train_data, k = k, kernel = "rectangular")
  train_predictions <- fitted(knn_model)
  
  # Calculate training misclassification error
  conf_matrix_train <- table(Actual = train_data$label, Predicted = train_predictions)
  train_correct <- sum(diag(conf_matrix_train))
  train_total <- sum(conf_matrix_train)
  train_errors[k] <- 1 - train_correct / train_total
  
  # Validate on validation data
  validation_predictions <- fitted(kknn(label ~ ., train = train_data, test = validation_data, k = k, kernel = "rectangular"))
  conf_matrix_validation <- table(Actual = validation_data$label, Predicted = validation_predictions)
  validation_correct <- sum(diag(conf_matrix_validation))
  validation_total <- sum(conf_matrix_validation)
  validation_errors[k] <- 1 - validation_correct / validation_total
}
# 
# # Plot training and validation errors with improvements
# plot(k_values, train_errors, type = "b", col = "blue", pch = 19, 
#      ylim = c(0, max(c(train_errors, validation_errors)) * 1.1), 
#      xlab = "Number of Neighbors (K)", 
#      ylab = "Misclassification Error", 
#      main = "Training and Validation Errors vs. K in KNN")
# 
# lines(k_values, validation_errors, type = "b", col = "red", pch = 19)
# 
# # Add gridlines
# abline(h = seq(0, max(c(train_errors, validation_errors)), by = 0.01), col = "gray", lty = 2)
# abline(v = seq(1, 30, by = 1), col = "gray", lty = 2)
# 
# # Highlight optimal K
# optimal_k <- which.min(validation_errors)
# points(optimal_k, validation_errors[optimal_k], col = "darkgreen", pch = 19, cex = 1.5)
# text(optimal_k, validation_errors[optimal_k], labels = paste("Optimal K =", optimal_k), pos = 3)
# 
# # Add legend
# legend("topright", legend = c("Training Error", "Validation Error"), 
#        col = c("blue", "red"), pch = 19, lty = 1, bty = "n")
# 


# Train the model with optimal K
optimal_k <- 3
knn_model_optimal <- kknn(label ~ ., train = train_data, test = test_data, k = optimal_k, kernel = "rectangular")

# Predictions and confusion matrix
test_predictions <- fitted(knn_model_optimal)
conf_matrix_test <- table(Actual = test_data$label, Predicted = test_predictions)

# Calculate test misclassification error
test_total <- sum(conf_matrix_test)
test_correct <- sum(diag(conf_matrix_test))
test_misclassification_error <- 1 - test_correct / test_total

# cat("Test Misclassification Error for K =", optimal_k, ":", test_misclassification_error, "\n")



# Initialize vector to store cross-entropy errors
cross_entropy_errors <- numeric(length(k_values))

# Loop through each value of K
for (k in k_values) {
  # Train KNN model and get predicted probabilities for validation data
  knn_model <- kknn(label ~ ., train = train_data, test = validation_data, k = k, kernel = "rectangular")
  predicted_probs <- knn_model$prob  # Predicted probabilities
  
  # Ensure probabilities are numeric (to avoid issues)
  predicted_probs <- as.matrix(predicted_probs)
  
  # Actual labels for the validation data (one-hot encoded)
  actual_labels <- model.matrix(~ label - 1, data = validation_data)
  
  # Add a small constant to avoid log(0) issues
  predicted_probs <- predicted_probs + 1e-15
  
  # Compute cross-entropy loss
  cross_entropy_loss <- -sum(actual_labels * log(predicted_probs)) / nrow(validation_data)
  
  # Store the cross-entropy loss
  cross_entropy_errors[k] <- cross_entropy_loss
}

# Plot cross-entropy errors for validation data
# plot(k_values, cross_entropy_errors, type = "b", col = "blue", pch = 19,
#      xlab = "Number of Neighbors (K)", ylab = "Cross-Entropy Error",
#      main = "Validation Cross-Entropy Error vs. K")

# Highlight optimal K
# optimal_k_cross_entropy <- which.min(cross_entropy_errors)
# optimal_cross_entropy_value <- cross_entropy_errors[optimal_k_cross_entropy]
# 
# # Mark the optimal K and cross-entropy value
# points(optimal_k_cross_entropy, optimal_cross_entropy_value, col = "darkgreen", pch = 19, cex = 1.5)
# text(optimal_k_cross_entropy, optimal_cross_entropy_value,
#      labels = paste("Optimal K =", optimal_k_cross_entropy, "\nError =", round(optimal_cross_entropy_value, 4)),
#      pos = 3)
# 
# # Add gridlines for better visualization
# abline(h = seq(0, max(cross_entropy_errors), by = 0.1), col = "gray", lty = 2)
# abline(v = seq(1, 30, by = 1), col = "gray", lty = 2)

```

-   **Training Misclassification Error**: `r train_misclassification_error`
-   **Test Misclassification Error**: `r test_misclassification_error`

**Confusion Matrix: Test Data**

```{r, echo=FALSE, results='asis'}
# Capture the confusion matrix as a character vector
test_matrix <- capture.output(conf_matrix_test)

# Print it in smaller font using LaTeX \small
cat("\\small\n\\begin{verbatim}\n")
cat(paste(test_matrix, collapse = "\n"))
cat("\n\\end{verbatim}\n\\normalsize\n")
```

**Confusion Matrix: Training Data**

```{r, echo=FALSE, results='asis'}
# Capture the confusion matrix as a character vector
train_matrix <- capture.output(conf_matrix_train)

# Print it in smaller font using LaTeX \small
cat("\\small\n\\begin{verbatim}\n")
cat(paste(train_matrix, collapse = "\n"))
cat("\n\\end{verbatim}\n\\normalsize\n")
```

## Summary of Prediction Quality
### Key Results:
- **Training Error**: 4.36%
- **Test Error**: 4.62%
- Strong generalization across datasets.

### Class-Specific Observations:
- **Best Performance**: Digit `0` (perfect classification).
- **Strong Performance**: Digits `1`, `2`, `6`, `7` (minimal errors).
- **Challenging Cases**:
  - Digits `3`, `4`, `5`: Overlap with similar digits.
  - Digits `8`, `9`: High confusion with `1` and `3`.

------------------------------------------------------------------------

**Visualizations of Easiest and Hardest Cases for Digit '8'**

```{r, echo=FALSE,  figures-side1, fig.show="hold", out.width="33.3%"}
par(mar = c(4, 4, .1, .1))

# Plot the easiest cases (Row 1)
heatmap(easiest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", 
         xlab = "", ylab = "", main = "Easiest Case 1")
heatmap(easiest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", 
         xlab = "", ylab = "", main = "Easiest Case 2")
```

```{r, echo=FALSE, figures-side, fig.show="hold", out.width="33.3%"}
# Add caption for Row 1
par(mar = c(0, 4, 1, 1)) # Ensure space for text on the left

# Plot the hardest cases (Row 2)
heatmap(hardest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", 
         xlab = "", ylab = "", main = "Hardest Case 1")
heatmap(hardest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", 
         xlab = "", ylab = "", main = "Hardest Case 2")
heatmap(hardest_matrices[[3]], Colv = NA, Rowv = NA, scale = "none", 
         xlab = "", ylab = "", main = "Hardest Case 3")

layout(1)
par(mar = c(5, 4, 4, 2)) # Reset to default
```
The hardest 3 cases are visually ambiguous and challenging to identify as "8," likely due to overlapping features or incomplete digit representation. In contrast, the easiest 2 cases are visually distinct and clearly recognizable as "8," making them straightforward to classify.

```{r, echo=FALSE,  figures3, fig.show="hold", out.width="80%"}
# Initialize vectors to store errors
k_values <- 1:30  # Values of K to test
train_errors <- numeric(length(k_values))  # Training errors
validation_errors <- numeric(length(k_values))  # Validation errors

# Loop through each value of K
for (k in k_values) {
  # Train KNN model on training data
  knn_model <- kknn(label ~ ., train = train_data, test = train_data, k = k, kernel = "rectangular")
  train_predictions <- fitted(knn_model)
  
  # Calculate training misclassification error
  conf_matrix_train <- table(Actual = train_data$label, Predicted = train_predictions)
  train_correct <- sum(diag(conf_matrix_train))
  train_total <- sum(conf_matrix_train)
  train_errors[k] <- 1 - train_correct / train_total
  
  # Validate on validation data
  validation_predictions <- fitted(kknn(label ~ ., train = train_data, test = validation_data, k = k, kernel = "rectangular"))
  conf_matrix_validation <- table(Actual = validation_data$label, Predicted = validation_predictions)
  validation_correct <- sum(diag(conf_matrix_validation))
  validation_total <- sum(conf_matrix_validation)
  validation_errors[k] <- 1 - validation_correct / validation_total
}

# Plot training and validation errors with improvements
plot(k_values, train_errors, type = "b", col = "blue", pch = 19, 
     ylim = c(0, max(c(train_errors, validation_errors)) * 1.1), 
     xlab = "Number of Neighbors (K)", 
     ylab = "Misclassification Error", 
     main = "Training and Validation Errors vs. K in KNN")

lines(k_values, validation_errors, type = "b", col = "red", pch = 19)

# Add gridlines
abline(h = seq(0, max(c(train_errors, validation_errors)), by = 0.01), col = "gray", lty = 2)
abline(v = seq(1, 30, by = 1), col = "gray", lty = 2)

# Highlight optimal K
optimal_k <- which.min(validation_errors)
points(optimal_k, validation_errors[optimal_k], col = "darkgreen", pch = 19, cex = 1.5)
text(optimal_k, validation_errors[optimal_k], labels = paste("Optimal K =", optimal_k), pos = 3)

# Add legend
legend("topright", legend = c("Training Error", "Validation Error"), 
       col = c("blue", "red"), pch = 19, lty = 1, bty = "n")
```


- **Model Complexity**:
  - As \( K \) increases, the model becomes less complex because it considers more neighbors for classification. This effectively smooths the decision boundaries.
  - For **small \( K \)** (e.g., \( K = 1 \)), the model is highly complex as it relies only on the nearest neighbor, leading to potential overfitting.

- **Effect on Errors**:
  - **Training Error**:
    - For small \( K \), the training error is very low (almost 0) due to overfitting.
    - As \( K \) increases, the training error gradually rises because the model becomes less sensitive to individual training points.
  - **Validation Error**:
    - Validation error initially decreases as \( K \) increases, reducing overfitting.
    - Beyond a certain point, validation error starts increasing due to underfitting when the model becomes overly smoothed.

---

## Optimal \( K \) According to the Plot

- From the plot, the **optimal \( K \)** corresponds to the minimum validation error.
- **Optimal \( K = 3 \)** in this case, as the validation error is lowest at this point.

## Comparison of Test, Training, and Validation Errors

The test misclassification error for \( K = 3 \) is **0.0325**, which is slightly higher than the validation error (**0.0282**) and significantly higher than the training error (**0.0094**). This indicates that the model generalizes well to unseen data while maintaining a low error on the training data.


```{r, echo=FALSE,  figures4, fig.show="hold", out.width="80%"}
# Initialize vector to store cross-entropy errors
cross_entropy_errors <- numeric(length(k_values))

# Loop through each value of K
for (k in k_values) {
  # Train KNN model and get predicted probabilities for validation data
  knn_model <- kknn(label ~ ., train = train_data, test = validation_data, k = k, kernel = "rectangular")
  predicted_probs <- knn_model$prob  # Predicted probabilities
  
  # Ensure probabilities are numeric (to avoid issues)
  predicted_probs <- as.matrix(predicted_probs)
  
  # Actual labels for the validation data (one-hot encoded)
  actual_labels <- model.matrix(~ label - 1, data = validation_data)
  
  # Add a small constant to avoid log(0) issues
  predicted_probs <- predicted_probs + 1e-15
  
  # Compute cross-entropy loss
  cross_entropy_loss <- -sum(actual_labels * log(predicted_probs)) / nrow(validation_data)
  
  # Store the cross-entropy loss
  cross_entropy_errors[k] <- cross_entropy_loss
}

# Plot cross-entropy errors for validation data
plot(k_values, cross_entropy_errors, type = "b", col = "blue", pch = 19,
     xlab = "Number of Neighbors (K)", ylab = "Cross-Entropy Error",
     main = "Validation Cross-Entropy Error vs. K")

# Highlight optimal K
optimal_k_cross_entropy <- which.min(cross_entropy_errors)
optimal_cross_entropy_value <- cross_entropy_errors[optimal_k_cross_entropy]

# Mark the optimal K and cross-entropy value
points(optimal_k_cross_entropy, optimal_cross_entropy_value, col = "darkgreen", pch = 19, cex = 1.5)
text(optimal_k_cross_entropy, optimal_cross_entropy_value, 
     labels = paste("Optimal K =", optimal_k_cross_entropy, "\nError =", round(optimal_cross_entropy_value, 4)), 
     pos = 3)
```

## Optimal \( K \) Value and Suitability of Cross-Entropy Error

- **Optimal \( K \)**:
  - From the graph, the **optimal \( K \)** is **19**, as the validation cross-entropy error is minimized at this point (**Error = 0.1456**).

- **Why Cross-Entropy is More Suitable**:
  - Cross-entropy considers the confidence of predictions by penalizing incorrect or low-probability predictions more heavily.
  - Unlike misclassification error, which treats all incorrect predictions equally, cross-entropy error differentiates between predictions with varying levels of certainty.
  - In a multinomial distribution (as in this case with multiple classes), cross-entropy provides a probabilistic perspective, which is essential for understanding model confidence and calibration, making it more informative for fine-tuning models.

# Assignment 2: Linear regression and ridge regression


```{r, echo=FALSE}
scale_features <- function(data) {
  data_scaled <- data
  data_scaled[ , -which(names(data) == "motor_UPDRS")] <- scale(data[ , -which(names(data) == "motor_UPDRS")])
  return(data_scaled)
}

# Define the log-likelihood function
Loglikelihood <- function(theta, sigma, X, y) {
  n <- nrow(X)
  y_pred <- X %*% theta
  residual_sum_squares <- sum((y - y_pred)^2)
  
  log_likelihood <- - (n / 2) * log(2 * pi) - (n / 2) * log(sigma^2) - 
    (1 / (2 * sigma^2)) * residual_sum_squares
  return(log_likelihood)
}

# Define Ridge log-likelihood function
Ridge <- function(theta, sigma, X, y, lambda) {
  log_likelihood <- Loglikelihood(theta, sigma, X, y)
  ridge_penalty <- lambda * sum(theta[-1]^2)  # Exclude intercept
  return(log_likelihood - ridge_penalty)
}

# Define Ridge optimization function
RidgeOpt <- function(X, y, lambda) {
  initial_theta <- rep(0, ncol(X))  # Initialize theta to zeros
  initial_sigma <- 1               # Initial guess for sigma
  initial_values <- c(initial_theta, initial_sigma)
  
  objective_function <- function(params) {
    theta <- params[1:(length(params) - 1)]
    sigma <- params[length(params)]
    return(-Ridge(theta, sigma, X, y, lambda))
  }
  
  # Minimize the negative Ridge log-likelihood
  result <- optim(par = initial_values, fn = objective_function, method = "BFGS")
  
  optimized_theta <- result$par[1:(length(result$par) - 1)]
  optimized_sigma <- result$par[length(result$par)]
  
  return(list(theta = optimized_theta, sigma = optimized_sigma, value = result$value))
}

# Function to calculate degrees of freedom for Ridge regression
DF <- function(X, lambda) {
  p <- ncol(X)
  I <- diag(p)
  XtX <- t(X) %*% X
  ridge_matrix <- solve(XtX + lambda * I)
  H <- X %*% ridge_matrix %*% t(X)
  return(sum(diag(H)))
}


#Load and Preprocess Data
# Load dataset
file_path <- "data/parkinsons.csv"
parkinson_data <- read.csv(file_path)

# Split into training and test sets
set.seed(123)
n <- nrow(parkinson_data)
train_indices <- sample(1:n, size = 0.6 * n)
train_data <- parkinson_data[train_indices, ]
test_data <- parkinson_data[-train_indices, ]

# Scale features
train_data_scaled <- scale_features(train_data)
test_data_scaled <- scale_features(test_data)


#Train Linear Model
# Fit linear model
linear_model <- lm(motor_UPDRS ~ ., data = train_data_scaled)
# Predictions
train_predictions <- predict(linear_model, train_data_scaled)
test_predictions <- predict(linear_model, test_data_scaled)

# Calculate MSE
train_mse <- mean((train_data_scaled$motor_UPDRS - train_predictions)^2)
test_mse <- mean((test_data_scaled$motor_UPDRS - test_predictions)^2)

# Display MSE
# <!-- cat("Train MSE:", train_mse, "\n") -->
# <!-- cat("Test MSE:", test_mse, "\n") -->



#Ridge Regression
# Prepare data for Ridge regression
X <- as.matrix(cbind(1, train_data_scaled[, -which(names(train_data_scaled) == "motor_UPDRS")]))
y <- train_data_scaled$motor_UPDRS

# Initial log-likelihood
theta_initial <- coef(linear_model)
sigma_initial <- summary(linear_model)$sigma
log_likelihood_value <- Loglikelihood(theta_initial, sigma_initial, X, y)
# cat("Log-Likelihood (Linear Model):", log_likelihood_value, "\n")

# Ridge optimization with 位 = 0.001
lambda <- 0.001
ridge_optimization_result <- RidgeOpt(X, y, lambda)

#Degrees of Freedom Analysis
# Degrees of Freedom for Ridge regression
lambda <- 10
degrees_of_freedom <- DF(X, lambda)

# Compute predictions and MSE
predict_ridge <- function(X, y, theta) {
  y_pred <- X %*% theta
  mse <- mean((y - y_pred)^2)
  return(list(predictions = y_pred, mse = mse))
}

# Perform Ridge regression and compute metrics for given lambda values
lambda_values <- c(1, 100, 1000)
results <- list()

for (lambda in lambda_values) {
  # Optimize Ridge parameters
  ridge_result <- RidgeOpt(X, y, lambda)
  
  # Extract optimized parameters
  optimized_theta <- ridge_result$theta
  optimized_sigma <- ridge_result$sigma
  
  # Predict on training and test data
  train_result <- predict_ridge(X, y, optimized_theta)
  test_result <- predict_ridge(as.matrix(cbind(1, test_data_scaled[, -which(names(test_data_scaled) == "motor_UPDRS")])), 
                               test_data_scaled$motor_UPDRS, optimized_theta)
  
  # Compute Degrees of Freedom
  degrees_of_freedom <- DF(X, lambda)
  
  # Store results
  results[[as.character(lambda)]] <- list(
    lambda = lambda,
    optimized_theta = optimized_theta,
    optimized_sigma = optimized_sigma,
    train_mse = train_result$mse,
    test_mse = test_result$mse,
    degrees_of_freedom = degrees_of_freedom
  )
}

# } 
```

-   **Training MSE**: `r train_mse`
-   **Test MSE**: `r test_mse`

The linear regression model shows that age, sex, total_UPDRS, Jitter.Abs., Shimmer.APQ5, and Shimmer.APQ11 are significant predictors of motor_UPDRS based on their low p-values (< 0.05). The total_UPDRS variable has the strongest influence, while others contribute moderately. The training MSE and test MSE indicate the model's accuracy on respective datasets.

## Optimal theta parameters for different lambda values

-   **lambda = 1**

```{r echo=FALSE}
lambda <- 1
ridge_optimization_result <- RidgeOpt(X, y, lambda)
```

-   **Optimized Theta (coefficients):**: `r ridge_optimization_result$theta`
-   **lambda = 100**

```{r echo=FALSE}
lambda <- 100
ridge_optimization_result <- RidgeOpt(X, y, lambda)
```

-   **Optimized Theta (coefficients):**: `r ridge_optimization_result$theta`
-   **lambda = 1000**

```{r echo=FALSE}
lambda <- 1
ridge_optimization_result <- RidgeOpt(X, y, lambda)
```

-   **Optimized Theta (coefficients):**: `r ridge_optimization_result$theta`

```{r echo=FALSE, results='asis'}
# Prepare a function to compute predictions and MSE
predict_ridge <- function(X, y, theta) {
  y_pred <- X %*% theta
  mse <- mean((y - y_pred)^2)
  return(list(predictions = y_pred, mse = mse))
}

# Perform Ridge regression and compute metrics for given lambda values
lambda_values <- c(1, 100, 1000)
results <- data.frame(
  Lambda = numeric(),
  Train_MSE = numeric(),
  Test_MSE = numeric(),
  Degrees_of_Freedom = numeric(),
  Optimized_Sigma = numeric()
)

for (lambda in lambda_values) {
  # Optimize Ridge parameters
  ridge_result <- RidgeOpt(X, y, lambda)
  
  # Extract optimized parameters
  optimized_theta <- ridge_result$theta
  optimized_sigma <- ridge_result$sigma
  
  # Predict on training and test data
  train_result <- predict_ridge(X, y, optimized_theta)
  test_result <- predict_ridge(as.matrix(cbind(1, test_data_scaled[, -which(names(test_data_scaled) == "motor_UPDRS")])), 
                               test_data_scaled$motor_UPDRS, optimized_theta)
  
  # Compute Degrees of Freedom
  degrees_of_freedom <- DF(X, lambda)
  
  # Append results to the data frame
  results <- rbind(
    results,
    data.frame(
      Lambda = lambda,
      Train_MSE = train_result$mse,
      Test_MSE = test_result$mse,
      Degrees_of_Freedom = degrees_of_freedom,
      Optimized_Sigma = optimized_sigma
    )
  )
}

# Display results as a table using knitr::kable
knitr::kable(results, format = "pipe", align = "c", caption = "Ridge Regression Results for Different Lambda Values")
```

From the table above, the choice of the penalty parameter **位** significantly affects the model's performance and complexity. Here's the detailed analysis:

- **lambda = 1**:
  - Achieves the lowest **Train MSE** (6.085893) and **Test MSE** (6.560836).
  - Maintains a **Degrees of Freedom (DF)** of 19.84347, meaning more predictors contribute effectively to the model.
  - Indicates a balance between model complexity and generalization.

- **lambda = 100**:
  - Results in higher **Train MSE** (25.88616) and **Test MSE** (26.94751) compared to **lambda = 1**.
  - Reduces **DF** to 15.59068, indicating a stronger penalty and increased regularization.
  - May lead to underfitting as the model becomes overly simplified.

- **lambda = 1000**:
  - Has the highest **Train MSE** (61.36093) and **Test MSE** (62.8066), demonstrating significant underfitting.
  - Reduces **DF** further to 10.04924, suggesting excessive shrinkage of predictor coefficients.
  - Results in a simpler but less accurate model.


Based on the results and analysis:

- **lambda = 1** is the most appropriate penalty parameter. It minimizes both **Train MSE** and **Test MSE**, while maintaining a high **Degrees of Freedom**, ensuring that the model is neither over-regularized nor under-regularized.
- As **lambda** increases, the model becomes overly regularized, leading to underfitting (higher MSE and lower DF).


# Appendix

## Code for Asssignment 1
```{r appendix_Asssignment_1, echo=TRUE, eval=FALSE}
#Load and Preprocess Data
# Load dataset
file_path <- "data/optdigits.csv"
optdigits_data <- read.csv(file_path)
colnames(optdigits_data) <- c(paste0("pixel_", 1:64), "label")
optdigits_data$label <- as.factor(optdigits_data$label)

# Split into training and test sets
set.seed(123)
n <- nrow(optdigits_data)
# Generate indices
train_indices <- sample(1:n, size = 0.5 * n)  # 50% for training
remaining_indices <- setdiff(1:n, train_indices)  # Remaining 50%
test_indices <- sample(remaining_indices, size = 0.25 * n)  # 25% for testing
validation_indices <- setdiff(remaining_indices, test_indices)  # Remaining 25% for validation

# Subset the data
train_data <- optdigits_data[train_indices, ]
test_data <- optdigits_data[test_indices, ]
validation_data <- optdigits_data[validation_indices, ]

# Train the KNN classifier
knn_model <- kknn(label ~ ., train = train_data, test = train_data, k = 30, kernel = "rectangular")


# Predictions and performance
train_predictions <- fitted(knn_model)
conf_matrix_train <- table(Actual = train_data$label, Predicted = train_predictions)
print("Confusion Matrix (Training):")
print(conf_matrix_train)

# Predict on test data
test_predictions <- fitted(kknn(label ~ ., train = train_data, test = test_data, k = 30, kernel = "rectangular"))
conf_matrix_test <- table(Actual = test_data$label, Predicted = test_predictions)
print("Confusion Matrix (Test):")
print(conf_matrix_test)


# Misclassification error for training data
train_total <- sum(conf_matrix_train)  # Total instances in training data
train_correct <- sum(diag(conf_matrix_train))  # Correctly classified instances (sum of diagonal)
train_misclassification_error <- 1 - train_correct / train_total

cat("Training Misclassification Error:", train_misclassification_error, "\n")

# Misclassification error for test data
test_total <- sum(conf_matrix_test)  # Total instances in training data
test_correct <- sum(diag(conf_matrix_test))  # Correctly classified instances (sum of diagonal)
test_misclassification_error <- 1 - test_correct / test_total

cat("Test Misclassification Error:", test_misclassification_error, "\n")


# Get predicted probabilities for training data
train_probabilities <- knn_model$prob  # Matrix of probabilities
train_actual <- train_data$label  # Actual labels for training data

# Extract rows corresponding to digit "8" in the training data
digit_8_indices <- which(train_actual == "8")
digit_8_probs <- train_probabilities[digit_8_indices, ]  # Probabilities for digit "8"

digit_8_predicted_probs <- digit_8_probs[, "8"]

easiest_indices <- digit_8_indices[order(digit_8_predicted_probs, decreasing = TRUE)[1:2]]
hardest_indices <- digit_8_indices[order(digit_8_predicted_probs, decreasing = FALSE)[1:3]]

# Extract feature data (pixels) for easiest and hardest cases
easiest_features <- train_data[easiest_indices, 1:64]  # Features for easiest cases
hardest_features <- train_data[hardest_indices, 1:64]  # Features for hardest cases

# Reshape into 8x8 matrices
# Reshape and store as numeric matrices
easiest_matrices <- lapply(1:nrow(easiest_features), function(i) {
  matrix(as.numeric(easiest_features[i, ]), nrow = 8, byrow = TRUE)
})

hardest_matrices <- lapply(1:nrow(hardest_features), function(i) {
  matrix(as.numeric(hardest_features[i, ]), nrow = 8, byrow = TRUE)
})

# Visualize easiest cases
heatmap(easiest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", main = paste("Easiest Case", 1), xlab = "", ylab = "")
heatmap(easiest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", main = paste("Easiest Case", 2), xlab = "", ylab = "")

# Visualize hardest cases
heatmap(hardest_matrices[[1]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", 1), xlab = "", ylab = "")
heatmap(hardest_matrices[[2]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", 2), xlab = "", ylab = "")
heatmap(hardest_matrices[[3]], Colv = NA, Rowv = NA, scale = "none", main = paste("Hardest Case", 3), xlab = "", ylab = "")



# Initialize vectors to store errors
k_values <- 1:30  # Values of K to test
train_errors <- numeric(length(k_values))  # Training errors
validation_errors <- numeric(length(k_values))  # Validation errors

# Loop through each value of K
for (k in k_values) {
  # Train KNN model on training data
  knn_model <- kknn(label ~ ., train = train_data, test = train_data, k = k, kernel = "rectangular")
  train_predictions <- fitted(knn_model)
  
  # Calculate training misclassification error
  conf_matrix_train <- table(Actual = train_data$label, Predicted = train_predictions)
  train_correct <- sum(diag(conf_matrix_train))
  train_total <- sum(conf_matrix_train)
  train_errors[k] <- 1 - train_correct / train_total
  
  # Validate on validation data
  validation_predictions <- fitted(kknn(label ~ ., train = train_data, test = validation_data, k = k, kernel = "rectangular"))
  conf_matrix_validation <- table(Actual = validation_data$label, Predicted = validation_predictions)
  validation_correct <- sum(diag(conf_matrix_validation))
  validation_total <- sum(conf_matrix_validation)
  validation_errors[k] <- 1 - validation_correct / validation_total
  
}
# 
# Plot training and validation errors
plot(k_values, train_errors, type = "b", col = "blue", pch = 19,
     ylim = c(0, max(c(train_errors, validation_errors)) * 1.1),
     xlab = "Number of Neighbors (K)",
     ylab = "Misclassification Error",
     main = "Training and Validation Errors vs. K in KNN")

lines(k_values, validation_errors, type = "b", col = "red", pch = 19)

# Add gridlines
abline(h = seq(0, max(c(train_errors, validation_errors)), by = 0.01), col = "gray", lty = 2)
abline(v = seq(1, 30, by = 1), col = "gray", lty = 2)

# Highlight optimal K
optimal_k <- which.min(validation_errors)
points(optimal_k, validation_errors[optimal_k], col = "darkgreen", pch = 19, cex = 1.5)
text(optimal_k, validation_errors[optimal_k], labels = paste("Optimal K =", optimal_k), pos = 3)

# Add legend
legend("topright", legend = c("Training Error", "Validation Error"),
       col = c("blue", "red"), pch = 19, lty = 1, bty = "n")



# Train the model with optimal K
optimal_k <- 3
knn_model_optimal <- kknn(label ~ ., train = train_data, test = test_data, k = optimal_k, kernel = "rectangular")

# Predictions and confusion matrix
test_predictions <- fitted(knn_model_optimal)
conf_matrix_test <- table(Actual = test_data$label, Predicted = test_predictions)

# Calculate test misclassification error
test_total <- sum(conf_matrix_test)
test_correct <- sum(diag(conf_matrix_test))
test_misclassification_error <- 1 - test_correct / test_total

cat("Test Misclassification Error for K =", optimal_k, ":", test_misclassification_error, "\n")

#Plot cross-entropy errors for validation data
plot(k_values, cross_entropy_errors, type = "b", col = "blue", pch = 19,
     xlab = "Number of Neighbors (K)", ylab = "Cross-Entropy Error",
     main = "Validation Cross-Entropy Error vs. K")

#Highlight optimal K
optimal_k_cross_entropy <- which.min(cross_entropy_errors)
optimal_cross_entropy_value <- cross_entropy_errors[optimal_k_cross_entropy]

# Mark the optimal K and cross-entropy value
points(optimal_k_cross_entropy, optimal_cross_entropy_value, col = "darkgreen", pch = 19, cex = 1.5)
text(optimal_k_cross_entropy, optimal_cross_entropy_value,
     labels = paste("Optimal K =", optimal_k_cross_entropy, "\nError =", round(optimal_cross_entropy_value, 4)),
     pos = 3)

# Add gridlines for better visualization
abline(h = seq(0, max(cross_entropy_errors), by = 0.1), col = "gray", lty = 2)
abline(v = seq(1, 30, by = 1), col = "gray", lty = 2)
```

## Code for Asssignment 2
```{r appendix_Asssignment_2, echo=TRUE, eval=FALSE}
scale_features <- function(data) {
  data_scaled <- data
  data_scaled[ , -which(names(data) == "motor_UPDRS")] <- scale(data[ , -which(names(data) == "motor_UPDRS")])
  return(data_scaled)
}

# Define the Loglikelihood function
Loglikelihood <- function(theta, sigma, X, y) {
  n <- nrow(X)
  y_pred <- X %*% theta
  residual_sum_squares <- sum((y - y_pred)^2)
  
  log_likelihood <- - (n / 2) * log(2 * pi) - (n / 2) * log(sigma^2) - 
    (1 / (2 * sigma^2)) * residual_sum_squares
  return(log_likelihood)
}

# Define Ridge function
Ridge <- function(theta, sigma, X, y, lambda) {
  log_likelihood <- Loglikelihood(theta, sigma, X, y)
  ridge_penalty <- lambda * sum(theta[-1]^2)  # Exclude intercept
  return(log_likelihood - ridge_penalty)
}

# Define Ridge optimization function
RidgeOpt <- function(X, y, lambda) {
  initial_theta <- rep(0, ncol(X))  # Initialize theta to zeros
  initial_sigma <- 1               # Initial guess for sigma
  initial_values <- c(initial_theta, initial_sigma)
  
  objective_function <- function(params) {
    theta <- params[1:(length(params) - 1)]
    sigma <- params[length(params)]
    return(-Ridge(theta, sigma, X, y, lambda))
  }
  
  # Minimize the negative Ridge log-likelihood
  result <- optim(par = initial_values, fn = objective_function, method = "BFGS")
  
  optimized_theta <- result$par[1:(length(result$par) - 1)]
  optimized_sigma <- result$par[length(result$par)]
  
  return(list(theta = optimized_theta, sigma = optimized_sigma, value = result$value))
}

# Function to calculate degrees of freedom for Ridge regression
DF <- function(X, lambda) {
  p <- ncol(X)
  I <- diag(p)
  XtX <- t(X) %*% X
  ridge_matrix <- solve(XtX + lambda * I)
  H <- X %*% ridge_matrix %*% t(X)
  return(sum(diag(H)))
}


#Load and Preprocess Data
# Load dataset
file_path <- "data/parkinsons.csv"
parkinson_data <- read.csv(file_path)

# Split into training and test sets
set.seed(123)
n <- nrow(parkinson_data)
train_indices <- sample(1:n, size = 0.6 * n)
train_data <- parkinson_data[train_indices, ]
test_data <- parkinson_data[-train_indices, ]

# Scale features
train_data_scaled <- scale_features(train_data)
test_data_scaled <- scale_features(test_data)


#Train Linear Model
# Fit linear model
linear_model <- lm(motor_UPDRS ~ ., data = train_data_scaled)
# Predictions
train_predictions <- predict(linear_model, train_data_scaled)
test_predictions <- predict(linear_model, test_data_scaled)

# Calculate MSE
train_mse <- mean((train_data_scaled$motor_UPDRS - train_predictions)^2)
test_mse <- mean((test_data_scaled$motor_UPDRS - test_predictions)^2)

# Display MSE
cat("Train MSE:", train_mse, "\n") 
cat("Test MSE:", test_mse, "\n") 



#Ridge Regression
# Prepare data for Ridge regression
X <- as.matrix(cbind(1, train_data_scaled[, -which(names(train_data_scaled) == "motor_UPDRS")]))
y <- train_data_scaled$motor_UPDRS

# Initial log-likelihood
theta_initial <- coef(linear_model)
sigma_initial <- summary(linear_model)$sigma
log_likelihood_value <- Loglikelihood(theta_initial, sigma_initial, X, y)
cat("Log-Likelihood (Linear Model):", log_likelihood_value, "\n")

# Ridge optimization with 位 = 0.001
lambda <- 0.001
ridge_optimization_result <- RidgeOpt(X, y, lambda)
cat("Optimized Theta (coefficients):\n", ridge_optimization_result$theta, "\n")
cat("Optimized Sigma (residual standard error):\n", ridge_optimization_result$sigma, "\n")
cat("Minimized Negative Log-Likelihood:\n", ridge_optimization_result$value, "\n")


#Degrees of Freedom Analysis
# Degrees of Freedom for Ridge regression
lambda <- 10
degrees_of_freedom <- DF(X, lambda)
#cat("Degrees of Freedom for Ridge Regression (位 =", lambda, "):", degrees_of_freedom, "\n")

# Compute predictions and MSE
predict_ridge <- function(X, y, theta) {
  y_pred <- X %*% theta
  mse <- mean((y - y_pred)^2)
  return(list(predictions = y_pred, mse = mse))
}

# Perform Ridge regression and compute metrics for given lambda values


#Display results
for (lambda in names(results)) {
result <- results[[lambda]]
cat("Lambda =", result$lambda, "\n")
cat("Train MSE:", result$train_mse, "\n")
cat("Test MSE:", result$test_mse, "\n")
cat("Degrees of Freedom:", result$degrees_of_freedom, "\n")
cat("Optimized Sigma:", result$optimized_sigma, "\n")
cat("\n")
}

```

## Code for Assignment 3
```{r}
set.seed(12345)
library(ggplot2)
df <- read.csv('data/pima-indians-diabetes.csv')

colnames(df) <- c('pregnant',
               'plasma',
               'blood_pressure',
               'triceps',
               'insulin',
               'bmi',
               'diabetes_pedigree',
               'age',
               'diabetes')

df$diabetes <- df$diabetes == 1

df |> ggplot() +
      geom_point(aes(x = age, y = plasma, color = diabetes)) +
      labs(x = "Age", y = "Plasma glucose concentration", color = "Has diabetes?")

df$diabetes

m = glm(diabetes ~ age + plasma, data=train, family="binomial")

df$predicted_prob <- predict(m, newdata = df, type = "response")
classification_threshold <- 0.5
df$prediction <- df$predicted_prob > classification_threshold

missclassification_rate <- sum(df$diabetes != df$prediction) / nrow(df)

# 0 = k1 + k2 * age + k3 * plasma
k1 <- summary(m)$coefficients[1]
k2 <- summary(m)$coefficients[2]
k3 <- summary(m)$coefficients[3]

# plasma = (-k1 -k2 * age) / k3
plasma_function <- function(x)(-k1 -k2 * x) / k3

df |> ggplot() +
      geom_point(aes(x = age, y = plasma, color = prediction)) +
      geom_function(fun = plasma_function) +
      labs(x = "Age", y = "Plasma glucose concentration", color = "Predicted to have diabetes?")

p <- 0.2
df$prediction02 <- df$predicted_prob > p
k4 <- log (p / (1-p))
plasma_function <- function(x)( k4 -k1 -k2 * x) / k3

df |> ggplot() +
      geom_point(aes(x = age, y = plasma, color = prediction02)) +
      geom_function(fun = plasma_function) +
      labs(x = "Age", y = "Plasma glucose concentration", color = "Predicted to have diabetes?")

p <- 0.8
df$prediction08 <- df$predicted_prob > p
k4 <- log (p / (1-p))
plasma_function <- function(x)( k4 -k1 -k2 * x) / k3


df |> ggplot() +
      geom_point(aes(x = age, y = plasma, color = prediction08)) +
      geom_function(fun = plasma_function) +
      labs(x = "Age", y = "Plasma glucose concentration", color = "Predicted to have diabetes?")


df$z1 <- df$plasma ^ 4
df$z2 <- df$plasma ^ 3 * df$age
df$z3 <- df$plasma ^ 2 * df$age ^ 2
df$z4 <- df$plasma  * df$age ^ 3
df$z5 <- df$age ^ 4

m2 = glm(diabetes ~ age + plasma + z1 + z2 + z3 + z4 + z5,
         data=df, family="binomial")

df$predicted_prob2 <- predict(m2, newdata = df, type = "response")
classification_threshold <- 0.5
df$prediction2 <- df$predicted_prob2 > classification_threshold

# 0 = k1 + k2 * age + k3 * plasma
k1 <- summary(m)$coefficients[1]
k2 <- summary(m)$coefficients[2]
k3 <- summary(m)$coefficients[3]

# plasma = (-k1 -k2 * age) / k3
plasma_function <- function(x)(-k1 -k2 * x) / k3

df |> ggplot() +
      geom_point(aes(x = age, y = plasma, color = prediction)) +
      geom_function(fun = plasma_function) +
      labs(x = "Age", y = "Plasma glucose concentration", color = "Predicted to have diabetes?")


```

